A comprehensive analysis of natural sequence variation within the <i>Arabidopsis thaliana</i> nuclear auxin signaling pathway using an accessible web application

### 2018-03-08 manuscript/manuscript.Rmd
Initial edits, stopped at "Identification of new Loss-of-Function"

next we substitute our deffinition of $n_{!i}$

@@

Auxin acts by binding to receptors that in turn target co-repressors for degradation. Auxin receptors (six in the model plant Arabidopsis thaliana) evolved through gene duplication and diversification early in the history of vascular plants @Parry_Complex_2009. The Nemhauser lab has recently shown that the rate of co-repressor degradation is determined by the identity of both the receptor and co-repressor @Havens_Synthetic_2012, and that this rate sets the pace of development3. 
Although it is unknown whether different receptors play different roles at the organismal level, a number of studies have pointed out differences in the ways the six different receptors in A. thaliana differ in their expression. The expression of AFB5 is induced in the presence of auxin, making this a circadian-regulated gene @Covington_Circadian_2007. It also appears AFB3 is more highly expressed in the roots in the presence of nitrate, allowing increased lateral root formation @Vidal_Nitrate_2010. Although TIR1/AFBs are expressed ubiquitously in A. thaliana tissues, TIR1, AFB2, and AFB3 have been shown to accumulate in the shoot and root meristems and leaf tissues, with slightly different expression patterns in TIR1 @Parry_Complex_2009. AFB1 has been found to have a mutation drastically reducing its ability to assemble into an SCF complex, contributing to it’s increased accumulation, as well as preventing it from degrading Aux/IAAs. Higher order TIR1/AFB mutants show AFB1 to have inconclusive organismal function. However, AFB1 is shown to be able to accumulate ubiquitously in all tissues @Yu_Untethering_2015. When it comes to binding auxins, AFB4 and AFB5 have been shown to preferentially and functionally bind the synthetic auxin picloram, while others have a preference for naturally occurring auxin 3-IAA @Prigge_Arabidopsis_2016. Interestingly, the ways in which TIR1/AFBs are able to bind and mark Aux/IAAs for degradation are variable. TIR1 and AFB2 have slightly different speed at which they perform this function, with AFB2 being generally quicker, as shown by our lab’s yeast degradation system @Havens_Synthetic_2012. Finally, higher order TIR1/AFB mutants show AFB1 to have inconclusive organismal function.
  Using the Natural Variation Webtool developed in our lab, I will characterize the natural variation in the TIR1/AFB family of auxin receptors by looking at the genomes of 1001 accessions of A. thaliana. This webtool allows me to	quantify and compare the nucleotide diversity of different genes in this family, as well as find relevant potentially function-altering variants. 
Using its diversity statistic function has shown me TIR1 and AFB1 both seem to be low diversity, hinting at their likely functional importance. AFB3 and 4 appear to be higher diversity, while their sisters are more conserved. This matches our current understanding of AFB3 and 4 as being minor players in the auxin signaling pathway, but brings up questions about the function of AFB1, which is inconclusive. Using this webtool, the AFB1 sequence can be aligned to the highly characterized TIR1 AA sequence showing locations of known non-synonymous variants. Although the most important functional regions do not contain variants in AFB1, there is a low diversity mutant, only found in the Can-1 accession, which is in the highly conserved TIR1/AFB oligomerization domain. Because variants eliciting significant changes in function often are not selected for, the fact that this variant is extremely low diversity holds promise that it is associated with a strong auxin signaling phenotype. I will order and characterize this accession in order to identify potential auxin signaling phenotypes. This would support that AFB1 functions through oligomerization with other TIR1/AFBs.
  The AFB4 and AFB4 receptors are known to have an extension in their N-terminal domains. Using the alignment feature on this webtool has indicated this region to be the highest in diversity in these proteins, containing lots of missense variants (Fig. 4). This suggests that this extension does not play an important functional role in these proteins.

@@

@Parry_Complex_2009
1.	Parry G, Calderon-Villalobos LI, Prigge M, et al. Complex regulation of the TIR1/AFB family of auxin receptors. Proceedings of the National Academy of Sciences of the United States of America. 2009;106(52):22540-22545. doi:10.1073/pnas.0911967106.
@Havens_Synthetic_2012
2.	Kyle A. Havens, et al. A Synthetic Approach Reveals Extensive Tunability of Auxin Signaling. Plant Physiology Sep 2012, 160 (1) 135-142; doi: 10.1104/pp.112.202184
@Guseman_Auxin-induced_2015
3.	Guseman JM, et al. Auxin-induced degradation dynamics set the pace for lateral root development. Development (Cambridge, England). 2015;142(5):905-909. doi:10.1242/dev.117234.
@Covington_Circadian_2007
8.	Covington MF, Harmer SL. The Circadian Clock Regulates Auxin Signaling and Responses in Arabidopsis . Weigel D, ed. PLoS Biology. 2007;5(8):e222. doi:10.1371/journal.pbio.0050222.
@Vidal_Nitrate_2010
9.	Vidal EA, Araus V, Lu C, et al. Nitrate-responsive miR393/AFB3 regulatory module controls root system architecture in Arabidopsis thaliana. Proceedings of the National Academy of Sciences of the United States of America. 2010;107(9):4477-4482. doi:10.1073/pnas.0909571107.
@Prigge_Arabidopsis_2016
10.	Prigge MJ, Greenham K, Zhang Y, et al. The Arabidopsis Auxin Receptor F-Box Proteins AFB4 and AFB5 Are Required for Response to the Synthetic Auxin Picloram. G3: Genes|Genomes|Genetics. 2016;6(5):1383-1390. doi:10.1534/g3.115.025585.
@Yu_Untethering_2015
11.	Yu H, Zhang Y, Moss BL, et al. Untethering the TIR1 auxin receptor from the SCF complex increases its stability and inhibits auxin response. Nature plants. 2015;1(3):14030. doi:10.1038/nplants.2014.30.
>>>>>>> AFB
## Aux/IAA genes

###Introduction
The Aux/IAA proteins (IAAs) mediate both auxin perception and transcriptional co-repression of auxin responsive genes. Auxin perception is carried out along with TIR1/AFB proteins, which together act as a coreceptor to perceive intracellular auxin. Interaction of IAAs with auxin and TIR1/AFBs is conferred by the highly conserved 13 amino acid degron (DII) motif [@tan_mechanism_2007]. Stabilizing mutations within the degron can lead to dominant phenotypes [@berleth_auxin_2004; @yang_iaa1_2004]. Degradation of IAAs relieves their repression upon ARF transcription factors. IAA degradation rate is influenced both by the degron sequence and by rate motifs found outside the degron, such as the KR motif [@moss_rate_2015; @dreher_arabidopsis_2006]. IAA-mediated repression of ARFs depends upon an EAR domain that recruits co-repressors of the TOPLESS (TPL) family, and the PB1 domain that enables interaction with ARF PB1 domains [@nanao_structural_2014; @kagale_ear_2011]. The Arabidopsis Aux/IAA gene family has 29 members, most of which possess a canonical degron. Notable exceptions are the Auxin insensitive IAA20 and IAA30 that lack a degron sequence entirely, while the closely related IAA31 retains a semi-functional degron sequence with a very slow degradation rate [@dreher_arabidopsis_2006; @havens_synthetic_2012]. Using the Natural Variation Webtool, we found that for the canonical Aux/IAAs, no missense mutations were present in the KR motif or PB1 interaction faces, and very few missense mutations were observed in the EAR domain and degron (Supp. Fig. Alignment of canonical IAAs). However, as discussed below, there were several interesting missense mutations found in EAR domains and degrons. The majority of nonsynonymous mutations observed were in the variable regions between the conserved domains (Supp. Fig. Alignment of canonical IAAs).

###Sister Pair Diversification
Previous research has pointed to genetic redundancy within closely related pairs or families of IAA proteins [@overvoorde_functional_2005]. For example, the iaa8-1, iaa9-1 double mutant and the iaa5-1, iaa6-1, iaa19-1 triple mutant have wild-type phenotypes [@overvoorde_functional_2005]. Despite this, most Arabidopsis IAA sister pairs have one member exhibiting higher diversity while the other sister is more conserved (Figure IAA phylogeny heat map). For example, despite having only minor differences in sequence (61.4% sequence identity), the IAA6/19 sister pair has significant differences in expression patterns, protein abundances and functions, suggesting they have undergone functional specialization since their divergence [@winkler_variation_2017]. IAA19 was expressed at higher levels than IAA6 and showed less sequence divergence; IAA6 showed regions of increased divergence upstream of the degron and in the N-terminal half of the PB1 domain [@winkler_variation_2017]. Consistent with these results, the Natural Variation Webtool (Figure IAA phylogeny heat map) revealed a lower natural variation for IAA19 (ᴨN/ᴨS = 0.55) compared to IAA6 (ᴨN/ᴨS = 2.3), and also detected high natural variation diversity within the same regions of IAA6 as seen in Winkler et al. (Supp Fig IAA6 diversity plot). Within IAA Group B, there are four sister pairs--IAA10 and IAA11, IAA12/BDL and IAA13, IAA18 and IAA26/PAP1, and IAA20 and IAA30 (Fig X) [@remington_contrasting_2004]. Interestingly, among these sister pairs only IAA12 and 13 have noticeable differences in diversity. IAA10 and IAA11 show little difference in ℼN/ℼS values, with ℼN/ℼS of 0.80 and 0.67 respectively, implicating the possibility that IAA10 and 11 play redundant roles. In rice however, it has been shown that IAA11 gain of function mutants block lateral root initiation [@zhu_gain--function_2012]. The Arabidopsis EFP browser shows that IAA10 and 11 have almost identical expression patterns, with the exception of IAA10 being more highly expressed in the developing seed, and IAA11 being more highly expressed in root tissue during lateral root development. As discussed below, the Natural Variation Webtool provides a potentially expedited route to conducting experiments meant to elucidate the extent of redundancy between family members such as IAA10 and 11.

>>>>

next we substitute our definition of $n_{!i}$

@@

Auxin acts by binding to receptors that in turn target co-repressors for degradation. Auxin receptors (six in the model plant Arabidopsis thaliana) evolved through gene duplication and diversification early in the history of vascular plants [@Parry_Complex_2009]. The rate of co-repressor degradation is determined by the identity of both the receptor and co-repressor [@Havens_Synthetic_2012], and that this rate sets the pace of development [@Guseman_Auxin-induced_2015].

Although it is unknown whether different receptors play different roles at the organismal level, a number of studies have pointed out differences in the ways the six different receptors in A. thaliana differ in their expression. The expression of AFB5 is induced in the presence of auxin, making this a circadian-regulated gene [@Covington_Circadian_2007]. Also, AFB3 is more highly expressed in the roots in the presence of nitrate, allowing increased lateral root formation [@Vidal_Nitrate_2010]. Although TIR1/AFBs are expressed ubiquitously in A. thaliana tissues, TIR1, AFB2, and AFB3 have been shown to accumulate in the shoot and root meristems and leaf tissues, with slightly different expression patterns for TIR1 [@Parry_Complex_2009]. AFB1 has a mutation drastically reducing its ability to assemble into an SCF complex, contributing to its increased accumulation, as well as preventing it from degrading Aux/IAAs[@Yu_Untethering_2015]. Higher order TIR1/AFB mutants show AFB1 to have inconclusive organismal function. However, AFB1 is shown to be able to accumulate ubiquitously in all tissues.AFB4 and AFB5 have been shown to preferentially and functionally bind the synthetic auxin picloram, while others have a preference for naturally occurring auxin indole-3-acetic acid [@Prigge_Arabidopsis_2016]. Interestingly, the strength and rate with which TIR1/AFBs are able to bind and mark Aux/IAAs for degradation are variable [@Havens_Synthetic_2012]. TIR1 and AFB2 have slightly different speed at which they perform this function, with AFB2 being generally faster.

Examining the natural sequence variation across the AFB revealed that *TIR1* and *AFB1* both had very low nonsynonymous diversity, hinting at their likely functional importance. *AFB3* and *AFB4* had higher nonsynonymous diversity, while their sisters genes, *AFB2* and *AFB5* are more conserved. This matches our current understanding of *AFB3* and *AFB4* as being minor players in the auxin signaling pathway, but brings in to question the inconclusive role of *AFB1* in auxin signaling. Although most important functional regions do not contain variants in AFB1, there is a low diversity mutant, only found in the Can-1 accession, which is in the highly conserved TIR1/AFB oligomerization domain. Because variants eliciting significant changes in function often are not selected for, the fact that this variant is extremely low diversity may suggest that it is associated with a strong auxin signaling phenotype. Characterization of this accession may identify potential *AFB1* phenotypes and may also test the hypothesis that AFB1 functions through oligomerization with other TIR1/AFBs.
  The AFB4 and AFB5 receptors have an N-terminal extension prior to the F-box domains. This extesion had very high nonsynonymous diversity(Fig. 4), suggesting that this extension does not play an important functional role in these proteins.

@@

## Aux/IAA genes

###Introduction
The Aux/IAA proteins (IAAs) mediate both auxin perception and transcriptional co-repression of auxin responsive genes. Auxin perception is carried out along with TIR1/AFB proteins, which together with IAAs act as a coreceptor to bind and perceive intracellular auxin. Interaction of IAAs with auxin and TIR1/AFBs is conferred by the highly conserved 13 amino acid degron motif [@tan_mechanism_2007]. Formation of this auxin coreceptor complex triggers ubiquitination and subsequent degradation of IAAs. This degradation in turn relieves their repression upon ARF transcription factors.Stabilizing mutations within the degron can lead to dominant phenotypes [@berleth_auxin_2004; @yang_iaa1_2004]. The Arabidopsis Aux/IAA gene family has 29 members, most of which possess a canonical degron. Notable exceptions are the auxin-insensitive IAA20 and IAA30 that lack a degron sequence entirely, while the closely related IAA31 retains a semi-functional degron sequence with a very slow degradation rate [@dreher_arabidopsis_2006; @havens_synthetic_2012]. Additionally, IAA degradation rate is influenced by rate motifs found outside the degron, such as the KR motif [@moss_rate_2015; @dreher_arabidopsis_2006]. IAA-mediated repression of ARFs depends upon both an EAR domain, N-terminal of the degron, that recruits co-repressors of the TOPLESS (TPL) family, and a Phox and Bem1 (PB1) domain that enables interaction via heteroligomerization with ARF PB1 domains [@nanao_structural_2014; @kagale_ear_2011].

Within the available natural sequence variation, we found that for the canonical Aux/IAAs, no missense mutations were present in the KR motif or residues responsible for PB1-PB1 domain interactions (Supp. Fig. Alignment of canonical IAAs). The majority of nonsynonymous mutations observed were in the variable regions between the conserved domains. However, several missense polymorphisms of interest were identified in EAR domains and degrons of several canonical IAAs.

###Sister Pair Diversification
Previous research has ~~pointed to~~ *found evidence of* genetic redundancy within closely related pairs or groups of IAA proteins [@overvoorde_functional_2005]. For example, the iaa8-1, iaa9-1 double mutant and the iaa5-1, iaa6-1, iaa19-1 triple mutant have wild-type phenotypes [@overvoorde_functional_2005]. Despite this, most *Arabidopsis* IAA sister pairs have one member with high nonsynonymous diversity while the other sister is more conserved (Figure IAA phylogeny heat map). For example, despite having only minor differences in sequence (61.4% sequence identity), the IAA6/19 sister pair has significant differences in expression patterns, protein abundances and functions, suggesting they have undergone functional specialization since their divergence [@winkler_variation_2017]. In an interspecific comparison between *A. thaliana* abd *A. lyrata*, IAA19 is more conserved, while IAA6 showed regions of increased sequence divergence upstream of the degron and in the N-terminal half of the PB1 domain [@winkler_variation_2017]. Consistent with these results, our analysis of intraspecific natural variation revealed higher conservation for IAA19 (${\pi_N/\pi_S}$ = 0.55) compared to IAA6 (${\pi_N/\pi_S}$ = 2.3) (Figure IAA phylogeny heat map), and also detected high diversity within the same regions of IAA6 as seen in Winkler et al. (Supp Fig IAA6 diversity plot).

Within the Group B IAA genes, there are four sister pairs--IAA10/IAA11, IAA12/IAA13, IAA18/IAA26, and IAA20/IAA30 (Fig X) [@remington_contrasting_2004]. Interestingly, among these sister pairs only IAA12 and 13 have a noticeable difference in nonsynonymous diversity. The Arabidopsis EFP browser shows that IAA10 and 11 have almost identical expression patterns, with the exception of IAA10 being more highly expressed in the developing seed, and IAA11 being more highly expressed in root tissue during lateral root development. In our analysis, IAA10 and IAA11 both showed functional conservation (${\pi_N/\pi_S}$ of 0.80 and 0.67 respectively) in support of IAA10 and IAA11 playing redundant roles. Accessions and associated variants identified here provide a potentially expedited route to elucidating the extent of redundancy between family members such as IAA10 and 11.
